---
title: "README"
output:
 github_document:
 pandoc_args: ["--wrap=none"]
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)


# Install required packages if not already installed
if (!require("httr")) install.packages("httr", dependencies = TRUE)

# Set Kaggle credentials (Ensure your Kaggle API key is in ~/.kaggle/kaggle.json)
Sys.setenv(KAGGLE_CONFIG_DIR = "~/.kaggle")

if (!require("kableExtra")) install.packages("kableExtra", dependencies = TRUE)
if (!require("knitr")) install.packages("knitr", dependencies = TRUE)
library(knitr)

# Install required package if not installed
if (!require("psych")) install.packages("psych", dependencies = TRUE)

# Load the library
library(psych)

library(gt)
library(tidyr)
library(dplyr)
library(kableExtra)
library(fBasics)
```


## Importing the Dataset
```{r}
# Define the dataset to download
dataset <- "fratzcan/usa-house-prices"

# Create a system command to download the dataset
system(paste("kaggle datasets download -d", dataset, "--unzip"))
```

```{r}
# Define the URL for the dataset
dataset_url <- "https://www.kaggle.com/api/v1/datasets/download/fratzcan/usa-house-prices"

# Define the destination file path (you can change this)
destination_file <- "usa-house-prices.zip"

# Download the dataset
download.file(dataset_url, destfile = destination_file, mode = "wb")

# Extract the dataset
unzip(destination_file, exdir = "usa-house-prices")

# List extracted files
list.files("usa-house-prices")

#Import dataset
housing_data <- read.csv ("USA Housing Dataset.csv", stringsAsFactors = FALSE )

#getwd()
```



```{r, echo =FALSE}
head(housing_data)
summary(housing_data)
str(housing_data)

#install.packages("DT")
#library(DT)
#datatable(housing_data)

#View(housing_data)

#List column names 
names(housing_data)

```


## Understanding the Data
```{r, eval=TRUE, include=FALSE}



#When was the data acquired? 
mind<-min(housing_data$date, na.rm = TRUE)
maxd<-max(housing_data$date, na.rm = TRUE)


#ANSWER: > min(housing_data$date, na.rm = TRUE)
#[1] "2014-05-02 00:00:00"
#> max(housing_data$date, na.rm = TRUE)
#[1] "2014-07-10 00:00:00"

#Where was the data acquired? (geographical location)
table(housing_data$city)
city_table <-as.data.frame(table(housing_data$city))
colnames(city_table) <-c("City", "Frequency")
where <- city_table

#View(city_table)

```
### When was the data acquired?

The data was gathered over the period of time from May 2, 2014 and July 10, 2014. The data was uploaded to Kaggle. The date this data was first accessed for this analysis was February 12, 2025. 
The dataset has the min date of `r mind`

### Where was the data acquired?

The data was acquired from Zillow from several different cities in Washington State.  The data was then uploaded to Kaggle. Below is a table that displays the number of houses in each city in Washington:  `r knitr::kable(where)`

 

### How was the data acquired?

### What are the attributes of this dataset?
The dataset contains eighteen different attributes. This section presents the attributes that appear in the dataset. 
First, we specify what each attribute is with descriptions and their data type. Then, we provide a codebook that specifies a number with each individual attribute. These numbers will be useful in determining which attribute is correlated with the summary statistics displayed in the last table.  

```{r}
# Create a data frame with attribute names and descriptions
attributes_table <- data.frame(
  Attribute = c("date", "price", "bedrooms", "bathrooms", "sqft_living",
                "sqft_lot", "floors", "waterfront", "view", "condition",  "sqft_above", "sqft_basement", "yr_built", "yr_renovated", "street", "city", "statezip", "country"),

  Description = c("Date when the house was sold",
                  "Sale price of the house",
                  "Number of bedrooms",
                  "Number of bathrooms",
                  "Living area in square feet",
                  "Lot area in square feet",
                  "Number of floors",
                  "Whether the house is waterfront (1: Yes, 0: No)",
                  "Number of times the house has been viewed",
                  "Condition of the house (1-5 scale)",
                  "Square footage of the house above ground",
                  "Square footage of the basement",
                  "Year the house was built",
                  "Year of renovation (0 if never renovated)",
                  "Street address of house",
                  "City of the house location",
                  "Zip code of the house location",
                  "Country of the house location")
)
```

```{r, eval=TRUE, include=FALSE}
# Print the table
print(attributes_table)
```

```{r, eval=TRUE, include=FALSE}

str(housing_data, list.len = ncol(housing_data))

```

```{r, eval=TRUE, include=FALSE}
#View(attributes_table)
#Finding the data types of each attribute 
print(typeof(housing_data$date))
print(typeof(housing_data$price))
print(typeof(housing_data$bedrooms))
print(typeof(housing_data$bathrooms))
print(typeof(housing_data$sqft_living))
print(typeof(housing_data$sqft_lot))
print(typeof(housing_data$floors))
print(typeof(housing_data$waterfront))
print(typeof(housing_data$view))
print(typeof(housing_data$condition))
print(typeof(housing_data$sqft_above))
print(typeof(housing_data$sqft_basement))
print(typeof(housing_data$yr_built))
print(typeof(housing_data$yr_renovated))
print(typeof(housing_data$statezip))
print(typeof(housing_data$country))

```



```{r}
# Add a column corresponding to each of the attributes data type 
attributes_table$Type = c ( "character", "double", "double", "double",
                "integer", "integer", "double", "integer", "integer", "integer",
                "integer", "integer", "integer", "integer",
                "character", "character", "character", "character")





# Generate a well-formatted table
styled_table <- kable(attributes_table, format = "html", caption = "📌 USA Housing Data") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"),
                full_width = F, position = "center") %>%
  column_spec(1, bold = TRUE, color = "black") %>%
  column_spec(2, italic = TRUE, color = "blue") %>%
  column_spec(3, width = "40em") %>%
  row_spec(0, bold = TRUE, color = "white", background = "black")

# Display the table
styled_table
```

```{r, eval=TRUE, include=FALSE}

#View(attributes_table)
```


## Summary Statistics

```{r, eval=TRUE, include=FALSE}
#PART 2
#Summary statistics for each attribute 

describe(housing_data)
summary(housing_data$price)
describe(housing_data$price)

# Select only numeric columns
numeric_data <- housing_data[sapply(housing_data, is.numeric)]

# Create a summary table, ensuring NAs are handled correctly
summary_table <- data.frame(
  Attribute = colnames(numeric_data),
  Min = sapply(numeric_data, function(x) min(x, na.rm = TRUE)),
  Max = sapply(numeric_data, function(x) max(x, na.rm = TRUE)),
  Mean = sapply(numeric_data, function(x) mean(x, na.rm = TRUE)),
  Median = sapply(numeric_data, function(x) median(x, na.rm = TRUE)),
  SD = sapply(numeric_data, function(x) sd(x, na.rm = TRUE))
)

# Print the corrected summary table
print(summary_table)
```




```{r, eval=TRUE, include=FALSE}
colnames(housing_data) <- c("Date", "Price", "# of Bedrooms", "# of Bathrooms", "Sqft of Living", "Sqft of Lot", "# of Floors", "Waterfront", "View", "Condition", "Sqft Above Ground", "Sqft of Basement", "Year Built", "Year Rennovated", "Street", "City", "Zipcode", "Country")
```

```{r}
Attribute <- colnames(housing_data)  # Extract column names
Code <- seq_along(Attribute)  # Create a sequence for the codes

# Create a proper data frame
df <- data.frame(Code, Attribute, stringsAsFactors = FALSE)

# Generate a well-formatted table
df %>% 
  kable(caption = "Attribute Codes for Housing Data", longtable = TRUE) %>%
  kable_styling(latex_options = c("hold_position", "repeat_header")) %>%
  row_spec(0, bold = TRUE) %>%
  column_spec(1, bold = TRUE)

```

```{r}

colnames (housing_data) = Code
housing_data3 = housing_data[-c(1:18)]
housing_data2 = housing_data[-c(1, 15:18)]
bstats = basicStats (housing_data2)[c("Mean", "Stdev", "Median", "Minimum", "Maximum",
"NAs"), ]
```

```{r, eval=TRUE, include=FALSE}
m<-matrix(1:ncol(bstats), 3)
for (i in 1:ncol(m)) {
cat(kable(bstats[, m[, i]], "latex", booktabs = TRUE, digits = 2, ), "\\newline")
}
```

```{r, eval=TRUE, include=FALSE}
print(kable(summary(bstats), format = "pipe", digits = 2))

summary_table <- as.data.frame(summary(bstats))
#gt(summary_table)
```

```{r, eval=TRUE, include=FALSE}
summary_stats <- bstats %>%
  summarise(across(everything(), list(
    Min = ~min(.x, na.rm = TRUE),
    Q1 = ~quantile(.x, 0.25, na.rm = TRUE),
    Median = ~median(.x, na.rm = TRUE),
    Mean = ~mean(.x, na.rm = TRUE),
    Q3 = ~quantile(.x, 0.75, na.rm = TRUE),
    Max = ~max(.x, na.rm = TRUE),
    SD = ~sd(.x, na.rm = TRUE)
  ), .names = "{.col}_{.fn}"))

summary_long <- summary_stats %>%
  pivot_longer(cols = everything(), names_to = c("Attribute", "Statistic"), names_sep = "_") %>%
  pivot_wider(names_from = "Attribute", values_from = "value") %>%
  mutate(across(where(is.numeric), ~round(.x, 2)))

print(summary_long)
```

```{r, eval=TRUE, include=FALSE}
summary_long <- summary_stats %>%
  pivot_longer(cols = everything(), names_to = c("Attribute", "Statistic"), names_sep = "_") %>%
  pivot_wider(names_from = "Attribute", values_from = "value") %>%
  mutate(across(where(is.numeric), ~round(.x, 2)))
print(summary_long)
```


```{r}
summary_long %>%
  kable(caption = "Summary Statistics for Housing Data", longtable = TRUE, align = "c") %>%
  kable_styling(latex_options = c("hold_position", "repeat_header")) %>%
  row_spec(0, bold = TRUE) %>%  # Bold column names (first row)
  row_spec(1:nrow(summary_long), font_size = 12) %>%  # Adjust font size for summary rows
  column_spec(1, bold = TRUE)   # Bold the first column (Statistic names)
```







