---
title: "README"
output:
 github_document:
 pandoc_args: ["--wrap=none"]
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)


# Install required packages if not already installed
if (!require("httr")) install.packages("httr", dependencies = TRUE)

# Set Kaggle credentials (Ensure your Kaggle API key is in ~/.kaggle/kaggle.json)
Sys.setenv(KAGGLE_CONFIG_DIR = "~/.kaggle")

if (!require("kableExtra")) install.packages("kableExtra", dependencies = TRUE)
if (!require("knitr")) install.packages("knitr", dependencies = TRUE)
library(knitr)

# Install required package if not installed
if (!require("psych")) install.packages("psych", dependencies = TRUE)

# Load the library
library(psych)

library(gt)
library(tidyr)
library(dplyr)
library(kableExtra)
library(fBasics)
```


## Importing the Dataset
```{r}
# Define the dataset to download
dataset <- "fratzcan/usa-house-prices"
```

```{r, eval=TRUE, include=FALSE}
# Create a system command to download the dataset
system(paste("kaggle datasets download -d", dataset, "--unzip"))
```

```{r}
# Define the URL for the dataset
dataset_url <- "https://www.kaggle.com/api/v1/datasets/download/fratzcan/usa-house-prices"

# Define the destination file path (you can change this)
destination_file <- "usa-house-prices.zip"

# Download the dataset
download.file(dataset_url, destfile = destination_file, mode = "wb")
```

```{r, eval=TRUE, include=FALSE}
# Extract the dataset
unzip(destination_file, exdir = "usa-house-prices")
```

```{r, eval=TRUE, include=FALSE}
# List extracted files
list.files("usa-house-prices")
```

```{r}
#Import dataset
housing_data <- read.csv ("USA Housing Dataset.csv", stringsAsFactors = FALSE )
```

```{r, eval=TRUE, include=FALSE}
#getwd()
```



```{r, echo =FALSE}
#head(housing_data)
#summary(housing_data)
#str(housing_data)

#install.packages("DT")
#library(DT)
#datatable(housing_data)

#View(housing_data)

#List column names 
#names(housing_data)

```


## Part 1
### Understanding the Housing Data
```{r, eval=TRUE, include=FALSE}



#When was the data acquired? 
mind<-min(housing_data$date, na.rm = TRUE)
maxd<-max(housing_data$date, na.rm = TRUE)


#ANSWER: > min(housing_data$date, na.rm = TRUE)
#[1] "2014-05-02 00:00:00"
#> max(housing_data$date, na.rm = TRUE)
#[1] "2014-07-10 00:00:00"

#Where was the data acquired? (geographical location)
table(housing_data$city)
city_table <-as.data.frame(table(housing_data$city))
colnames(city_table) <-c("City", "Frequency")
where <- city_table

#View(city_table)

```
**When was the data acquired?**

The data was gathered over the period of time from May 2, 2014 and July 10, 2014. The data was uploaded to Kaggle. The date this data was first accessed for this analysis was February 12, 2025. 


**Where was the data acquired?**

The data consists of housing prices from various cities in Washington, US. The cities are displayed in a frequency table:  `r knitr::kable(where)`


**How was the data acquired?**

The dataset was acquired from Kaggle. Kaggle does not provide any additional informaation about the source the data was collected from. There are no links to original publications or projects from which the dataset was derived. 


**What are the attributes of this dataset?**

The dataset contains eighteen different attributes. This section presents the attributes that appear in the dataset. 

First, we specify what each attribute is with descriptions and their data type (see Table 1). Then, we provide a codebook that specifies a number with each individual attribute. These numbers will be useful in determining which attribute is correlated with the summary statistics displayed in Table 2.  

```{r,eval=TRUE, include=FALSE}
# Create a data frame with attribute names and descriptions
attributes_table <- data.frame(
  Attribute = c("Date", "Price", "Bedrooms", "Bathrooms", "Sqft living",
                "Sqft Lot", "Floors", "Waterfront", "View", "Condition",  "Sqft Above", "Sqft Basement", "Year Built", "Year Rennovated", "Street", "City", "Statezip", "Country"),

  Description = c("Date when the house was sold",
                  "Sale price of the house",
                  "Number of bedrooms the house has",
                  "Number of bathrooms the house has",
                  "Living area in square feet",
                  "Lot area in square feet",
                  "Number of floors the house has",
                  "Whether the house is waterfront (1: Yes, 0: No)",
                  "Number of times the house has been viewed",
                  "Condition of the house (1-5 scale)",
                  "Square footage of the house above ground",
                  "Square footage of the basement",
                  "Year the house was built",
                  "Year of renovation (0 if never renovated)",
                  "Street address of house",
                  "City of the house location",
                  "Zip code of the house location",
                  "Country of the house location")
)
```

```{r, eval=TRUE, include=FALSE}
# Print the table
print(attributes_table)
```

```{r, eval=TRUE, include=FALSE}

str(housing_data, list.len = ncol(housing_data))

```



```{r, echo = FALSE}
get_mode <- function(x) {
  uniq_x <- unique(x)
  uniq_x[which.max(tabulate(match(x, uniq_x)))]
}

modes_selected<-lapply(housing_data, get_mode)

modes_table<-data.frame(Attribute = names(modes_selected), Mode = unlist(modes_selected))

print(modes_table)
```


```{r, eval=TRUE, include=FALSE}
#View(attributes_table)
#Finding the data types of each attribute 
print(typeof(housing_data$date))
print(typeof(housing_data$price))
print(typeof(housing_data$bedrooms))
print(typeof(housing_data$bathrooms))
print(typeof(housing_data$sqft_living))
print(typeof(housing_data$sqft_lot))
print(typeof(housing_data$floors))
print(typeof(housing_data$waterfront))
print(typeof(housing_data$view))
print(typeof(housing_data$condition))
print(typeof(housing_data$sqft_above))
print(typeof(housing_data$sqft_basement))
print(typeof(housing_data$yr_built))
print(typeof(housing_data$yr_renovated))
print(typeof(housing_data$statezip))
print(typeof(housing_data$country))

```



```{r, eval=TRUE, include=FALSE}
# Add a column corresponding to each of the attributes data type 
attributes_table$Type = c ( "character", "double", "double", "double",
                "integer", "integer", "double", "integer", "integer", "integer",
                "integer", "integer", "integer", "integer",
                "character", "character", "character", "character")





# Generate a well-formatted table
styled_table <- kable(attributes_table, format = "html", caption = "Table 1: 📌 USA Housing Data") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"),
                full_width = F, position = "center") %>%
  column_spec(1, bold = TRUE, color = "black") %>%
  column_spec(2, italic = TRUE, color = "blue") %>%
  column_spec(3, width = "40em") %>%
  row_spec(0, bold = TRUE, color = "white", background = "black")

# Display the table
styled_table
```

```{r, eval=TRUE, include=FALSE}

#View(attributes_table)
```


# Part 2
## Summary Statistics

```{r, eval=TRUE, include=FALSE}
#PART 2
#Summary statistics for each attribute 

describe(housing_data)
summary(housing_data$price)
#describe(housing_data$price)

# Select only numeric columns
numeric_data <- housing_data[sapply(housing_data, is.numeric)]
```

```{r}
# Create a summary table, ensuring NAs are handled correctly
summary_table <- data.frame(
  Attribute = colnames(numeric_data),
  Min = sapply(numeric_data, function(x) min(x, na.rm = TRUE)),
  Max = sapply(numeric_data, function(x) max(x, na.rm = TRUE)),
  Mean = sapply(numeric_data, function(x) mean(x, na.rm = TRUE)),
  Median = sapply(numeric_data, function(x) median(x, na.rm = TRUE)),
  SD = sapply(numeric_data, function(x) sd(x, na.rm = TRUE))
)
```
```{r, eval=TRUE, include=FALSE}
# Print the corrected summary table
print(summary_table)
```




```{r, eval=TRUE, include=FALSE}
colnames(housing_data) <- c("Date", "Price", "Number of Bedrooms", "Number of Bathrooms", "Sqft of Living", "Sqft of Lot", "Number of Floors", "Waterfront", "View", "Condition", "Sqft Above Ground", "Sqft of Basement", "Year Built", "Year Rennovated", "Street", "City", "Zipcode", "Country")
```

```{r}
Attribute <- c("Date of Sale", "Sale Price", "Number of Bedrooms", "Number of Bathrooms", 
               "Square Footage of Living Space", "Square Footage of Lot", "Number of Floors", 
               "Waterfront Property (1=Yes, 0=No)", "View Rating", "Condition Rating", 
               "Square Footage Above Ground", "Square Footage of Basement", 
               "Year Built", "Year Renovated", "Street Name", "City Name", "Zip Code", "Country")

# Generate a sequence for the codes
Code <- seq_along(Attribute)  

# Manually define the data types to align with the attribute descriptions
Type <- c("character", "double", "double", "double",
          "integer", "integer", "double", "integer", 
          "integer", "integer", "integer", "integer", 
          "integer", "integer", "character", "character",
          "character", "character")  

# Ensure that all three vectors have the same length
if (length(Code) == length(Attribute) && length(Attribute) == length(Type)) {
  # Create a proper data frame
  df <- data.frame(Code, Attribute, Type, stringsAsFactors = FALSE)
} else {
  stop("Mismatch in vector lengths! Check your Attribute and Type definitions.")
}

# Generate a well-formatted table
df %>% 
  kable(caption = "Attribute Codes for Housing Data", longtable = TRUE) %>%
  kable_styling(latex_options = c("hold_position", "repeat_header")) %>%
  row_spec(0, bold = TRUE) %>%
  column_spec(1, bold = TRUE) 
```


```{r}

colnames (housing_data) = Code
housing_data3 = housing_data[-c(1:18)]
housing_data2 = housing_data[-c(1, 15:18)]
bstats = basicStats (housing_data2)[c("Mean", "Stdev", "Median", "Minimum", "Maximum",
"NAs"), ]
```

```{r, eval=TRUE, include=FALSE}
m<-matrix(1:ncol(bstats), 3)
for (i in 1:ncol(m)) {
cat(kable(bstats[, m[, i]], "latex", booktabs = TRUE, digits = 2, ), "\\newline")
}
```


```{r, eval=TRUE, include=FALSE}
print(kable(summary(bstats), format = "pipe", digits = 2))

summary_table <- as.data.frame(summary(bstats))
#gt(summary_table)
```

```{r, eval=TRUE, include=FALSE}
summary_stats <- bstats %>%
  summarise(across(everything(), list(
    Min = ~min(.x, na.rm = TRUE),
    Q1 = ~quantile(.x, 0.25, na.rm = TRUE),
    Median = ~median(.x, na.rm = TRUE),
    Mean = ~mean(.x, na.rm = TRUE),
    Q3 = ~quantile(.x, 0.75, na.rm = TRUE),
    Max = ~max(.x, na.rm = TRUE),
    SD = ~sd(.x, na.rm = TRUE)
  ), .names = "{.col}_{.fn}"))

summary_long <- summary_stats %>%
  pivot_longer(cols = everything(), names_to = c("Attribute", "Statistic"), names_sep = "_") %>%
  pivot_wider(names_from = "Attribute", values_from = "value") %>%
  mutate(across(where(is.numeric), ~round(.x, 2)))

print(summary_long)
```

```{r, eval=TRUE, include=FALSE}
summary_long <- summary_stats %>%
  pivot_longer(cols = everything(), names_to = c("Attribute", "Statistic"), names_sep = "_") %>%
  pivot_wider(names_from = "Attribute", values_from = "value") %>%
  mutate(across(where(is.numeric), ~round(.x, 2)))
print(summary_long)
```


```{r}
summary_long %>%
  kable(caption = "Table 2: Summary Statistics for Housing Data", longtable = TRUE, align = "c") %>%
  kable_styling(latex_options = c("hold_position", "repeat_header")) %>%
  row_spec(0, bold = TRUE) %>%  # Bold column names (first row)
  row_spec(1:nrow(summary_long), font_size = 12) %>%  # Adjust font size for summary rows
  column_spec(1, bold = TRUE)   # Bold the first column (Statistic names)
```

```{r}
# Apply log1p (log(x + 1)) transformation to avoid log(0) issues
log_prices <- log1p(housing_data$`2`)

# Check the range of log-transformed prices
range_log_prices <- range(log_prices, na.rm = TRUE)


plot(housing_data$`3`, log_prices, 
     main="Figure 1. House Price vs Number of Bedrooms (Log-transformed)", 
     xlab = "Bedrooms", ylab = "Log of House Price + 1", 
     pch=19, cex = 1.5, col = rgb(0.2, 0.5, 0.8, 0.6), 
     xlim = c(min(housing_data$`3`), 
              max(housing_data$`3`)), 
     ylim = range_log_prices)
```

Figure 1 shows the relationship between the number of bedrooms each house has and the price of the house, scaled to the log (x+1) so it easier to visually understand and to eliminate the issue with values that were zero. It is clear that there is a higher concentration when the houses have three to six bedrooms by the cluster of points in the middle of the scatter plot, indicating the most expensive houses. Not many houses have zero or eight bedrooms, so the points appear more transparent (less concentrated). Figure 1 also depicts outliers at the bottom the graph, where the prices of the houses are significantly lower. 

```{r}
plot(housing_data$`4`, log_prices, 
     main="Figure 2. House Price vs Number of Bathrooms (Log-transformed)", 
     xlab = "Bathrooms", ylab = "Log of House Price + 1", 
     pch=19, cex = 1.5, col = rgb(0.2, 0.5, 0.8, 0.6), 
     xlim = c(min(housing_data$`4`), 
              max(housing_data$`4`)), 
     ylim = range_log_prices)
```

Figure 2 shows the relationship between the number of bathrooms the house has and the price of the house, scaled to the log(x+1). It seems like there the houses become more expensive as the number of bathrooms increase. There is a large concentration around 1.5 to 3.5 bathrooms, indicating the most houses within this range. There are outliers depicted at the bottom of the scatter plot which show that there are houses with lower prices than the majority of the dataset. Once we get 4.5 bathrooms, the points become more transparent, showcasing that the number of bathrooms is rare, even within the most expensive houses.

```{r}
hist(housing_data$`10`, 
     main= "Figure 3. Histogram of Housing Condition", 
     xlab= "Condition (1-5)", 
     ylab = "Frequency", 
     col = rgb(0.2, 0.5, 0.8))
```

Figure 3 shows the frequency of the condition of the houses in the dataset. The condition variable is shown on a 1-5 scale: 1 being the worst and 5 being the best. Here we can see that the houses in the dataset are mostly rated 3, with 4 being the next frequent value. By looking at this visualization, we can understand the general quality of houses in Washington, US. There are very few houses in poor condition. 

**Identifying Missing Values (NAs)**

In our code for generating the summary statistics table, we take into account the null values for the character/categorical attributes and make sure to exclude them from our summary statistic table. Our code numbers align with the numbers presented in our coded attribute table; there are five categorical attributes, meaning that there is no numeric data associated with them. We excluded these from the summary statistic table through our code. 



# Part 3
## Expanding Investment Knowledge 

**Why would this dataset be useful?**

This additional dataset consists of the all-transactions house price index for Washington, US, recorded quarterly, from 1975 to present. House price index (HPI) is an economic indicator that measures the change in average residential property prices over time. It shows how much house prices are rising or falling in a given region, usually calculated by tracking repeat sales of the same properties over time. It is considered a key tool for understanding housing market trends and affordability. 

This dataset would be useful because the original dataset consists of data from Washington, US only from May-July 2014, so it would be useful to compare the data with the HPI in Washington throughout the years to see whether housing prices in the state were relatively low or high at that time.
 

**How could it complement the data you're currently analyzing?** 

This dataset could complement the data we are currently analyzing to show the housing price index in the year 2014. We could compare the change in HPI from Quarter 2 and Quarter 3 to our own dataset, since our data was recorded from the months of May to August, which covers both quarters 2 and 3. This will help us analyze and compare the summary statistics of our data to the HPI of Washington in 2014. 

In addtion, we could use this dataset to compare the trends in housing prices over the years. The HPI shows the average original residential property prices have risen over time in Washington. It shows a decline following the 2008 recession, with the lowest peak displayed in 2012, then a gradual increase. The HPI had not yet hit the pre-recession high of 498.06 from 2007. It hit this high again in 2016 and has continued to increase since then. Our data is from 2014, so the data we collected falls in the recovering years of recession, showcasing unique trends. 

**Link to dataset**

[Additional Dataset](https://fred.stlouisfed.org/series/WASTHPI#)





